---
title: "Quality of Weight lifting exercises"
author: "Cherie"
date: "18/05/2015"
output: html_document
---

#Overview
In this project, the goal is to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants and predict whether they performed barbell lifts correctly and incorrectly in 5 different ways (classe). 

The data link is https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

This write up document describes the analysis. 

#Getting and cleaning data
The dataset was split in a training (60%) and testing (40%) set. Columns with a significant number (96%) of Missing values (NA) were removed and columns with significant Near-zero-values. Additionally, the X column was identified as a confounding variable and removed. The classes of the training and test sets were unified.
```{r, cache=TRUE, include=FALSE }
#loading data
data <- read.csv("pml-training.csv")
library(caret)
set.seed(125)
inTrain <- createDataPartition(y=data$classe, p=.6, list=F)
training <- data[inTrain,]
testing <- data[-inTrain,]

testCases <- read.csv("pml-testing.csv")
```

```{r, echo=TRUE, cache=TRUE}
dim(training); dim(testing)
```

```{r, include=FALSE, cache=TRUE}
# identifying and removing colums where NA values account for 96%
colNA <- colMeans(is.na(training))
training2<- training[, colNA < .96]
testing2<- testing[, colNA < .96]
testCases2 <- testCases[, colNA < .96]
#removing column 1 "X" which is a confounding variable 
training2 <- training2[, -1]
testing2 <- testing2[, -1]
testCases2 <- testCases2[,-1]
```

```{r, include=FALSE, cache=TRUE}
library(caret)
nzv <- nearZeroVar(training2)
training3 <- training2[,-nzv]; 
testing3 <- testing2[,-nzv]; 
testCases3 <- testCases2[,-nzv]
```

```{r, include=FALSE, cache=TRUE}
classTR <- class(training3)
class(testing3) <- classTR
class(testCases3) <- classTR
```

##Exploratory Data Analysis
The remaining data (58 variables) was reviewed with functions dim, str, summary. The distribution of the outcome variable was reviewed and featurePlots of the 58 variables were run.

```{r, echo=FALSE, fig.height=3, fig.width=4}
(library(lattice))
histogram(testing3$classe)
```

##Variable Importance and model selection
Relative importance is assessed using varImp function before and after NZV were removed. NZV proved to have little impact.
```{r, echo=TRUE, cache=TRUE }
RocImp_inclNZV <- filterVarImp(x=training2[, -ncol(training2)], y=training2$classe)
RocImp_exclNZV <- filterVarImp(x=training3[, -ncol(training3)], y=training3$classe)
summary(RocImp_exclNZV)
```

Three model were selected (with the accuracy statistic from the confusionMatrix alongside):
1. with all 58 remaining variables (91%)
2. included variables with importance > 70% (65% accuracy)
3. variables with importance > 60% (93% accuracy)

To improve processing speed parallel processing options were used to rund the model. The confusion matrix on the testing set is below. Method "gbm" is used and "PCA" pre-processing.
```{r, include=FALSE, cache=TRUE }
set.seed(125)
library(doMC)
registerDoMC(cores=5)
##All subsequent models are then run in parallel
gbmGrid <- expand.grid(interaction.depth=c(2,5,13), n.trees=c(150:2000), .shrinkage=c(0.001, .01))
fitControl <- trainControl(method="cv", number=5)
modelClass <- train(classe ~ accel_arm_x + magnet_arm_x +magnet_arm_y +roll_dumbbell + pitch_dumbbell + magnet_dumbbell_x + magnet_dumbbell_y +pitch_forearm + accel_forearm_x + magnet_forearm_x + cvtd_timestamp + num_window +roll_belt +accel_belt_z+magnet_belt_y+magnet_belt_z+roll_arm +pitch_arm+total_accel_arm+accel_arm_z+magnet_arm_z+roll_dumbbell+pitch_dumbbell+yaw_dumbbell+total_accel_dumbbell + accel_dumbbell_x + accel_dumbbell_y + accel_dumbbell_z + roll_forearm + pitch_forearm + total_accel_forearm + accel_forearm_x + magnet_forearm_x + magnet_forearm_y + magnet_forearm_z, data=training3, method="gbm", trControl=fitControl, preProc=c("pca"))
cM <- confusionMatrix(training3$classe, predict(modelClass, training3[,-58]))
```
The confusionMatrix below shows the results compared to the actual classe in the training set.
```{r, echo=FALSE}
cM
plot(modelClass)
```

The model was run on the testing set and the confusion Matrix of model, which is printed below, created.
```{r, echo=FALSE}
library(caret)
predClass <- predict(modelClass, testing3[,-58])
cMPred <- confusionMatrix(testing3$classe, predClass)
cMPred
```

The model was used to predict the classe of the 20 test cases which were submitted to Coursera.
```{r, echo=FALSE, cache=TRUE}
library(caret)
answers <- predict(modelClass, testCases3[,-58])
answers
```

20 Individual files were created using the code below which was provided.
pml_write_files = function(x){
+     n = length(x)
+     for(i in 1:n){
+         filename = paste0("problem_id_",i,".txt")
+         write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
+     }
+ }
pml_write_files(answers)


